---
title: "Kernel-Based Sparsity Diagnostic (KBSD)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kernel-Based Sparsity Diagnostic}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The KBSD package provides tools to diagnose sparsity and support issues using a kernel-based notion of effective data points (EDP).
Instead of relying purely on treatment density, this approach evaluates how much information is available about the outcome at a given point under different intervention schemes.

## Simulated Example

We begin with a simple simulated dataset.

```{r}
library(KBSD)
library(simcausal)

n <- 100

D.base <- DAG.empty() + 
  node("L", distr = "rbern", prob = 0.3) +
  node("A", distr = "rnorm", mean = 0.21 * L + 0.45, sd = 0.14)

Dset <- set.DAG(D.base)

Odat <- sim(DAG = Dset, n = n, rndseed = 1, verbose = FALSE)
Odat$ID <- NULL
```

### Intervention Scenarios

We define three intervention strategies:

```{r}
# Static intervention
Odat_int1 <- Odat
Odat_int1$A <- 1

# Small shift
Odat_int2 <- Odat
Odat_int2$A <- Odat_int2$A + 0.1

# Larger shift
Odat_int3 <- Odat
Odat_int3$A <- Odat_int3$A + 0.5
```

### Apply Diagnostic

```{r}
data_result <- kbsd(
  data = Odat,
  int_data_list = list(Odat_int1, Odat_int2, Odat_int3),
  disthalf_vec = c(L = 0.5, A = 0.1)
)

kbsd_plot(data_result)
```

## EFV Data: Single Time Point

We now apply the diagnostic to real data.

```{r}
library(CICI)

data(EFV)

efv_subset <- EFV[, c("log_age", "efv.0", "adherence.1")]
```

### Intervention Scenarios

```{r}
# Static intervention
efv_sub1 <- transform(efv_subset, efv.0 = 0)

# Modified treatment policy
efv_sub2 <- transform(efv_subset, efv.0 = efv.0 + 1)
```

### Apply Diagnostic

```{r}
data_result1 <- kbsd(
  data = efv_subset,
  int_data_list = list(efv_sub1, efv_sub2),
  disthalf_vec = c(log_age = 1, efv.0 = 0.5, adherence.1 = 0)
)

kbsd_plot(data_result1)
```

## EFV Data: Multiple Time Points

We now consider a longitudinal setting.

### Intervention Strategies

```{r}
# Strategy 1: static intervention
EFV_strat1 <- EFV
EFV_strat1[paste0("efv.", 0:4)] <- 3

# Strategy 2: modified treatment policy
EFV_strat2 <- EFV
EFV_strat2[paste0("efv.", 0:4)] <- 
  EFV_strat2[paste0("efv.", 0:4)] + 0.1
```

### Apply Diagnostic

```{r}
data_result2 <- kbsd(
  data = EFV,
  int_data_list = list(EFV_strat1, EFV_strat2),
  disthalf_vec = get_disthalf_vec(
    EFV,
    interventions = paste0("efv.", 0:4),
    remove = c("VL.4", "metabolic", "NRTI"),
    no_extrapolation = "sex"
  )
)

kbsd_plot(data_result2)
```

## Interpretation

The kernel-based sparsity diagnostic quantifies how much information is available for estimating counterfactual outcomes.

- High EDP values indicate strong local support in the data
- Low EDP values suggest extrapolation is needed at an unreasonable level and potential estimation instability
- Increasing the half-distance parameters increases EDP values by relaxing similarity requirements

## Summary

This vignette illustrated:

- How to apply the kernel-based sparsity diagnostic
- How to define intervention scenarios
- How to interpret effective data points

For further details, consult the function documentation:


?kbsd

?kbsd_plot

?get_disthalf_vec



